Transfer of Value Functions via Variational Methods

### 2018-05-18 NIPS 2018/main.tex
Reformatting

\begin{document}
% \nipsfinalcopy is no longer used

@@

Given the source tasks' weights $\mathcal{W}_s$, we model our estimated prior as a mixture of Gaussians with one equally weighted isotropic Gaussian centered at each weight: $p(\bm{w}) = \frac{1}{\abs{\mathcal{W}_s}}\sum_{\bm{w}_s \in \mathcal{W}_s}\mathcal{N}(\bm{w}|\bm{w}_s,\sigma_p^2\bm{I})$. This model resembles a kernel density estimator \cite{scott2015multivariate} and, due to its nonparametric nature, it allows capturing arbitrarily complex distributions. Consistently with the prior, we model our approximate posterior as a mixture of Gaussians. However, we allow a different number of components \todo{questo dettaglio si potrebbe spostare nella parte sperimentale o togliere del tutto} (typically much less than the prior's), and we adopt full covariances instead of only diagonals so that our posterior has the potential to match complex distributions with fewer components. Using $C$ components, our posterior is $q_{\bm{\xi}}(\bm{w}) = \frac{1}{C}\sum_{i=1}^C\mathcal{N}(\bm{w}|\bm{\mu}_i,\bm{\Sigma}_i)$, with variational parameters $\bm{\xi} = (\bm{\mu}_1,\dots,\bm{\mu}_C,\bm{\Sigma}_1,\dots,\bm{\Sigma}_C)$. Once again, we learn Cholesky factors instead of full covariances.

Although this new model has the potential to capture much more complicated distributions, it poses a significant complication: the KL divergence between two mixtures of Gaussians has no closed-form expression. To solve this issue, we rely on an upper bound to such quantity, so that negative ELBO still upper bounds the KL between the approximate and the exact posterior. Among the many upper bounds available in the literature, we adopt the one proposed in \cite{hershey2007approximating}, which we report here for the sake of completeness. We refer the reader to the original paper for the proof.
\begin{theorem}[\cite{hershey2007approximating}]\label{th:ukl}
Let $p = \sum_{i} c_i^{(p)} f_i^{(p)}$ and $q = \sum_{j} c_j^{(q)} f_j^{(q)}$ be two mixture of Gaussian distributions, where $f_i^{(p)} = \mathcal{N}(\bm{\mu}_i^{(p)},\bm{\Sigma}_i^{(p)})$ denotes the i-th component of $p$, $c_i^{(p)}$ denotes its weight, and similarly for $q$. Introduce two vectors $\chi^{(1)}$ and $\chi^{(2)}$ such that $c_i^{(p)} = \sum_j \chi^{(2)}_{j,i}$ and $c_j^{(q)} = \sum_i \chi^{(1)}_{i,j}$. Then:
\begin{equation}
KL(p||q) \leq KL(\chi^{(2)}||\chi^{(1)}) + \sum_{i,j} \chi^{(2)}_{j,i}KL(f_i^{(p)}||f_j^{(q)})
\end{equation}
%\begin{equation}
%\chi^{(2)}_{i,j} = \frac{c_j^{(q)}\chi^{(1)}_{j,i}}{\sum_{i'} \chi^{(1)}%_{j,i'}},\ \ \ \ \chi^{(1)}_{j,i} = \frac{c_i^{(p)}\chi^{(2)}_{i,j}e^{-%KL(f_i^{(p)}||f_j^{(q)})}}{\sum_{j'}\chi^{(2)}_{i,j'}e^{-KL(f_i^{(p)}||f_{j'}%^{(q)})}}
%\end{equation}
\end{theorem}
Our new algorithm replaces the KL with the above-mentioned upper bound. Each time we require its value, we have to recompute the parameters $\chi^{(1)}$ and $\chi^{(2)}$ that tighten the bound. As shown in \cite{hershey2007approximating}, we can use a simple fixed-point procedure for this purpose. Finally, both terms in the objective are now linear combinations of functions of the variational parameters of different components, and their gradients easily derive from the ones of the Gaussian case. We report the derivation in App. \ref{}.

\subsection{Optimizing the TD error} \label{sec:td}

From Sections \ref{sec:gvt} and \ref{sec:mgvt}, we know that differentiating the negative ELBO $\mathcal{L}$ w.r.t. $\bm{\xi}$ requires differentiating $\norm{B_{\bm{w}}}_D^2$ w.r.t. $\bm{w}$. Unfortunately, the TD error is well-known to be non-differentiable due to the presence of the max operator. This issue rarely is a problem since typical value-based algorithms are semi-gradient methods, i.e., they do not differentiate the targets (see, e.g., Chapter 11 of \cite{sutton1998reinforcement}). However, our transfer settings are quite different from common RL. In fact, our algorithm is likely to start from $Q$-functions that are very close to an optimum and aims only to adapt the weights in some direction of lower error (i.e., higher likelihood) so as to quickly converge to the solution of the target task. Unfortunately, this property does not hold for most semi-gradient algorithms. Even worse, many online RL algorithms combined with complex function approximators (e.g., DQNs) are well-known to be unstable, especially when approaching an optimum, and require many tricks and tuning to work well \cite{schaul2015prioritized,van2016deep}. This property is clearly undesirable in our case, as we only aim at adapting already good solutions. Thus, we consider using a residual gradient algorithm \cite{baird1995residual}. To differentiate the targets, we replace the optimal Bellman operator with the mellow Bellman operator introduced in \cite{asadi2017alternative}, which adopts a softened version of max called \textit{mellowmax}:
\begin{equation}
\mm_a Q_{\bm{w}}(s,a) = \frac{1}{\kappa} \log \frac{1}{|\mathcal{A}|}\sum_a e^{\kappa Q_{\bm{w}}(s,a)}
\end{equation}
where $\kappa$ is a hyperparameter and $|\mathcal{A}|$ is the number of actions. The mellow Bellman operator, which we denote as $\wt{T}$, has several appealing properties that make it suitable for our settings: (i) it converges to the maximum as $\kappa \rightarrow \infty$, (ii) it has a unique fixed point, and (iii) it is \textit{differentiable}. Denoting by $\wt{B}_{\bm{w}} = \wt{T}Q_{\bm{w}} - Q_{\bm{w}}$ the Bellman residual w.r.t. the mellow Bellman operator $\wt{T}$, we have that the corresponding TD error, $\norm{\wt{B}_{\bm{w}}}_D^2$, is now differentiable w.r.t. $\bm{w}$. Although residual algorithms have guaranteed convergence, they are typically much slower than their semi-gradient counterpart. \cite{baird1995residual} proposed to project the gradient in a direction that achieves higher learning speed, while preserving convergence. This projection is obtained by including a parameter $\psi\in[0,1]$ in the TD error gradient:
\begin{equation}
\nabla_{\bm{w}}\norm{\wt{B}_{\bm{w}}}_{D}^2 = \frac{2}{N}\sum_{i=1}^N b_i(\bm{w})\left( \gamma\psi\nabla_{\bm{w}}\mm_{a'}Q_{\bm{w}}(s_i',a') - \nabla_{\bm{w}}Q_{\bm{w}}(s_i,a_i)) \right),
\end{equation}
where $b_i(\bm{w}) = r_i + \gamma\mm_{a'}Q_{\bm{w}}(s_i',a') - Q_{\bm{w}}(s_i,a_i)$. Notice that $\psi$ trades-off between the semi-gradient ($\psi = 0$) and the full residual gradient ($\psi = 1$). A good criterion for choosing such parameter is to start with values close to zero (to have faster learning) and move to higher values when approaching the optimum (to guarantee convergence).

@@

Since the agent does not know the locations of the doors in advance and receives only very sparse feedback, it must efficiently explore the environment to figure out (i) their positions, and (ii) how to reach the goal. While this might be a complicated problem for plain RL, our transfer algorithm should be able to figure out the door positions quickly. In fact, notice that, although different, the optimal $Q$-functions for all tasks share some similarities. For example, once the agent has passed the last door before the goal, the $Q$-values are exactly the same in all tasks. The same does not hold for positions nearby the starting state. However, it is clear that there should be a preference over actions \textit{up} and \textit{right}, rather than \textit{down} and \textit{left} (which are worse in all tasks).\todo{Good candidate to be removed}

@@

To prove that our guesses are correct, we generate a set of $50$ source tasks for the three-room environment of Figure~\ref{f:rooms} by sampling both door locations uniformly in the allowed space, and solve all of them by directly minimizing the TD error as presented in Section~\ref{sec:td}. Then, we use our algorithms to transfer from $10$ source tasks sampled from the previously generated set. The average return over the last $50$ learning episodes as a function of the number of iterations is shown in Figure~\ref{f:two_doors}. Each curve is the result of $20$ independent runs, each one resampling the target and source tasks, with $95\%$ confidence intervals. Further details on the parameters adopted in this experiment are given in App. \ref{}. As expected, the no-transfer (NT) algorithm fails at learning the task in so few iterations due to the limited exploration provided by an $\epsilon$-greedy policy. On the other hand, all our algorithms achieve a significant speed-up and converge to the optimal performance in few iterations, with GVT being slightly slower. Interestingly, we notice that there is no advantage in adopting more than $1$ component for the posterior in MGVT. This result is intuitive since, as soon as the algorithm figures out which is the target task, all the components move towards the same region.

@@

Our approach is mostly related to \cite{lazaric2010bayesian}. Although we both assume the tasks to share similarities in their value functions, \cite{lazaric2010bayesian} consider only linear approximators and adopt a hierarchical Bayesian model of the corresponding weights' distribution, which is assumed Gaussian. On the other hand, our variational approximation allows for more general distribution families and can be combined with non-linear approximators. Furthermore, \cite{lazaric2010bayesian} propose a Dirichlet process model for the case where weights cluster into different classes, which relates to our mixture formulation and proves the importance of capturing more complicated task distributions again. In \cite{wilson2007multi}, the authors propose a hierarchical Bayesian model for the distribution over MDPs. unlike our approach and \cite{lazaric2010bayesian}, they consider a distribution over transition probabilities and rewards, rather than value functions. In the same spirit of our method, they consider a Thompson sampling-based procedure which, at each iteration, samples a new task from the posterior and solves it. However, \cite{wilson2007multi} consider only finite MDPs, which poses a severe limitation on the algorithm's applicability. On the contrary, our approach can handle high-dimensional tasks. In \cite{doshi2016hidden}, the authors consider a family of tasks whose dynamics are governed by some hidden parameters and use Gaussian processes (GPs) to model such dynamics across tasks. Recently, \cite{killian2017robust} extended this approach by replacing GPs with Bayesian neural networks, to obtain a more scalable approach.\todo{More comments about HiP-MDPs? Should we discuss more related works?}

@@

@@

For the implementation of the Mixture of Gaussian Variational Transfer, we use the upper bound on the KL divergence between two Mixtures of Gaussians, as in Theorem~\ref{th:ukl}, to obtain an upper bound on the negative ELBO in Equation~\eqref{eq:elbo}. Consider we have $C$ components for the posterior family $q_{\bm{\xi}}(\bm{w}) = \frac{1}{C}\sum_{i=1}^C\mathcal{N}(\bm{w}|\bm{\mu}_i,\bm{\Sigma}_i)$ and a prior distribution, constructed from the set of weights $\mathcal{W}_s = \left\lbrace \bm{w}_1,..., \bm{w}_{|\mathcal{W}_s|} \right\rbrace$ of the sources' optimal $Q$-functions, $p(\bm{w}) = \frac{1}{\abs{\mathcal{W}_s}}\sum_{j=1}^{|\mathcal{W}_s|}\mathcal{N}(\bm{w}|\bm{w}_j,\sigma_p^2\bm{I})$.

@@

>>>>

\begin{document}
\setlength{\abovedisplayskip}{1.5pt}
\setlength{\belowdisplayskip}{1.5pt} 
% \nipsfinalcopy is no longer used

@@

Given the source tasks' weights $\mathcal{W}_s$, we model our estimated prior as a mixture of Gaussians with one equally weighted isotropic Gaussian centered at each weight: $p(\bm{w}) = \frac{1}{\abs{\mathcal{W}_s}}\sum_{\bm{w}_s \in \mathcal{W}_s}\mathcal{N}(\bm{w}|\bm{w}_s,\sigma_p^2\bm{I})$. This model resembles a kernel density estimator \cite{scott2015multivariate} and, due to its nonparametric nature, it allows capturing arbitrarily complex distributions. Consistently with the prior, we model our approximate posterior as a mixture of Gaussians.  Using $C$ components, our posterior is $q_{\bm{\xi}}(\bm{w}) = \frac{1}{C}\sum_{i=1}^C\mathcal{N}(\bm{w}|\bm{\mu}_i,\bm{\Sigma}_i)$, with variational parameters $\bm{\xi} = (\bm{\mu}_1,\dots,\bm{\mu}_C,\bm{\Sigma}_1,\dots,\bm{\Sigma}_C)$. Once again, we learn Cholesky factors instead of full covariances. Finally, since the KL divergence between two mixtures of Gaussians has no closed-form expression, we rely on an upper bound to such quantity, so that negative ELBO still upper bounds the KL between the approximate and the exact posterior. Among the many upper bounds available in the literature, we adopt the one proposed in \cite{hershey2007approximating}, which we report in App. \ref{} for the sake of completeness.

\subsection{Optimizing the TD Error} \label{sec:td}

From Sections \ref{sec:gvt} and \ref{sec:mgvt}, we know that differentiating the negative ELBO $\mathcal{L}$ w.r.t. $\bm{\xi}$ requires differentiating $\norm{B_{\bm{w}}}_D^2$ w.r.t. $\bm{w}$. Unfortunately, the TD error is well-known to be non-differentiable due to the presence of the max operator. This issue rarely is a problem since typical value-based algorithms are semi-gradient methods, i.e., they do not differentiate the targets (see, e.g., Chapter 11 of \cite{sutton1998reinforcement}). However, our transfer settings are quite different from common RL. In fact, our algorithm is likely to start from $Q$-functions that are very close to an optimum and aims only to adapt the weights in some direction of lower error so as to quickly converge to the solution of the target task. Unfortunately, this property does not hold for most semi-gradient algorithms. Even worse, many online RL algorithms combined with complex function approximators (e.g., DQNs) are well-known to be unstable, especially when approaching an optimum, and require many tricks and tuning to work well \cite{schaul2015prioritized,van2016deep}. This property is clearly undesirable in our case, as we only aim at adapting already good solutions. Thus, we consider using a residual gradient algorithm \cite{baird1995residual}. To differentiate the targets, we replace the optimal Bellman operator with the mellow Bellman operator introduced in \cite{asadi2017alternative}, which adopts a softened version of max called \textit{mellowmax}:
\begin{equation}
\mm_a Q_{\bm{w}}(s,a) = \frac{1}{\kappa} \log \frac{1}{|\mathcal{A}|}\sum_a e^{\kappa Q_{\bm{w}}(s,a)}
\end{equation}
where $\kappa$ is a hyperparameter and $|\mathcal{A}|$ is the number of actions. The mellow Bellman operator, which we denote as $\wt{T}$, has several appealing properties that make it suitable for our settings: (i) it converges to the maximum as $\kappa \rightarrow \infty$, (ii) it has a unique fixed point, and (iii) it is \textit{differentiable}. Denoting by $\wt{B}_{\bm{w}} = \wt{T}Q_{\bm{w}} - Q_{\bm{w}}$ the Bellman residual w.r.t. the mellow Bellman operator $\wt{T}$, we have that the corresponding TD error, $\norm{\wt{B}_{\bm{w}}}_D^2$, is now differentiable w.r.t. $\bm{w}$. Further considerations on how to better optimize it are given in App. \ref{}.

@@

%Since the agent does not know the locations of the doors in advance and receives only very sparse feedback, it must efficiently explore the environment to figure out (i) their positions, and (ii) how to reach the goal. While this might be a complicated problem for plain RL, our transfer algorithm should be able to figure out the door positions quickly. In fact, notice that, although different, the optimal $Q$-functions for all tasks share some similarities. For example, once the agent has passed the last door before the goal, the $Q$-values are exactly the same in all tasks. The same does not hold for positions nearby the starting state. However, it is clear that there should be a preference over actions \textit{up} and \textit{right}, rather than \textit{down} and \textit{left} (which are worse in all tasks).\todo{Good candidate to be removed}

@@

We generate a set of $50$ source tasks for the three-room environment of Figure~\ref{f:rooms} by sampling both door locations uniformly in the allowed space, and solve all of them by directly minimizing the TD error as presented in Section~\ref{sec:td}. Then, we use our algorithms to transfer from $10$ source tasks sampled from the previously generated set. The average return over the last $50$ learning episodes as a function of the number of iterations is shown in Figure~\ref{f:two_doors}. Each curve is the result of $20$ independent runs, each one resampling the target and source tasks, with $95\%$ confidence intervals. Further details on the parameters adopted in this experiment are given in App. \ref{}. As expected, the no-transfer (NT) algorithm fails at learning the task in so few iterations due to the limited exploration provided by an $\epsilon$-greedy policy. On the other hand, all our algorithms achieve a significant speed-up and converge to the optimal performance in few iterations, with GVT being slightly slower. Interestingly, we notice that there is no advantage in adopting more than $1$ component for the posterior in MGVT. This result is intuitive since, as soon as the algorithm figures out which is the target task, all the components move towards the same region.

@@

Our approach is mostly related to \cite{lazaric2010bayesian}. Although we both assume the tasks to share similarities in their value functions, \cite{lazaric2010bayesian} consider only linear approximators and adopt a hierarchical Bayesian model of the corresponding weights' distribution, which is assumed Gaussian. On the other hand, our variational approximation allows for more general distribution families and can be combined with non-linear approximators. Furthermore, \cite{lazaric2010bayesian} propose a Dirichlet process model for the case where weights cluster into different classes, which relates to our mixture formulation and proves the importance of capturing more complicated task distributions again. In \cite{wilson2007multi}, the authors propose a hierarchical Bayesian model for the distribution over MDPs. unlike our approach and \cite{lazaric2010bayesian}, they consider a distribution over transition probabilities and rewards, rather than value functions. In the same spirit of our method, they consider a Thompson sampling-based procedure which, at each iteration, samples a new task from the posterior and solves it. However, \cite{wilson2007multi} consider only finite MDPs, which poses a severe limitation on the algorithm's applicability. On the contrary, our approach can handle high-dimensional tasks. In \cite{doshi2016hidden}, the authors consider a family of tasks whose dynamics are governed by some hidden parameters and use Gaussian processes (GPs) to model such dynamics across tasks. Recently, \cite{killian2017robust} extended this approach by replacing GPs with Bayesian neural networks, to obtain a more scalable approach.

@@

\subsection{Optimizing the TD Error}

Although residual algorithms have guaranteed convergence, they are typically much slower than their semi-gradient counterpart. \cite{baird1995residual} proposed to project the gradient in a direction that achieves higher learning speed, while preserving convergence. This projection is obtained by including a parameter $\psi\in[0,1]$ in the TD error gradient:
\begin{equation}
\nabla_{\bm{w}}\norm{\wt{B}_{\bm{w}}}_{D}^2 = \frac{2}{N}\sum_{i=1}^N b_i(\bm{w})\left( \gamma\psi\nabla_{\bm{w}}\mm_{a'}Q_{\bm{w}}(s_i',a') - \nabla_{\bm{w}}Q_{\bm{w}}(s_i,a_i)) \right),
\end{equation}
where $b_i(\bm{w}) = r_i + \gamma\mm_{a'}Q_{\bm{w}}(s_i',a') - Q_{\bm{w}}(s_i,a_i)$. Notice that $\psi$ trades-off between the semi-gradient ($\psi = 0$) and the full residual gradient ($\psi = 1$). A good criterion for choosing such parameter is to start with values close to zero (to have faster learning) and move to higher values when approaching the optimum (to guarantee convergence).

@@

As mentioned in the main paper, for the mixture version of Alg. \ref{alg:transfer} we rely on the upper bound on the KL divergence between two mixture of Gaussians presented in \cite{hershey2007approximating}. We report it here for the sake of completeness.

\begin{theorem}[\cite{hershey2007approximating}]\label{th:ukl}
Let $p = \sum_{i} c_i^{(p)} f_i^{(p)}$ and $q = \sum_{j} c_j^{(q)} f_j^{(q)}$ be two mixture of Gaussian distributions, where $f_i^{(p)} = \mathcal{N}(\bm{\mu}_i^{(p)},\bm{\Sigma}_i^{(p)})$ denotes the i-th component of $p$, $c_i^{(p)}$ denotes its weight, and similarly for $q$. Introduce two vectors $\chi^{(1)}$ and $\chi^{(2)}$ such that $c_i^{(p)} = \sum_j \chi^{(2)}_{j,i}$ and $c_j^{(q)} = \sum_i \chi^{(1)}_{i,j}$. Then:
\begin{equation}
KL(p||q) \leq KL(\chi^{(2)}||\chi^{(1)}) + \sum_{i,j} \chi^{(2)}_{j,i}KL(f_i^{(p)}||f_j^{(q)})
\end{equation}
%\begin{equation}
%\chi^{(2)}_{i,j} = \frac{c_j^{(q)}\chi^{(1)}_{j,i}}{\sum_{i'} \chi^{(1)}%_{j,i'}},\ \ \ \ \chi^{(1)}_{j,i} = \frac{c_i^{(p)}\chi^{(2)}_{i,j}e^{-%KL(f_i^{(p)}||f_j^{(q)})}}{\sum_{j'}\chi^{(2)}_{i,j'}e^{-KL(f_i^{(p)}||f_{j'}%^{(q)})}}
%\end{equation}
\end{theorem}
Our new algorithm replaces the KL with the above-mentioned upper bound. Each time we require its value, we have to recompute the parameters $\chi^{(1)}$ and $\chi^{(2)}$ that tighten the bound. As shown in \cite{hershey2007approximating}, we can use a simple fixed-point procedure for this purpose. Finally, both terms in the objective are now linear combinations of functions of the variational parameters of different components, and their gradients easily derive from the ones of the Gaussian case.

Consider we have $C$ components for the posterior family $q_{\bm{\xi}}(\bm{w}) = \frac{1}{C}\sum_{i=1}^C\mathcal{N}(\bm{w}|\bm{\mu}_i,\bm{\Sigma}_i)$ and a prior distribution, constructed from the set of weights $\mathcal{W}_s = \left\lbrace \bm{w}_1,..., \bm{w}_{|\mathcal{W}_s|} \right\rbrace$ of the sources' optimal $Q$-functions, $p(\bm{w}) = \frac{1}{\abs{\mathcal{W}_s}}\sum_{j=1}^{|\mathcal{W}_s|}\mathcal{N}(\bm{w}|\bm{w}_j,\sigma_p^2\bm{I})$. Then:

@@

\begin{figure*}[t]
\centering
    \includegraphics[trim=0.0cm 0cm 1.2cm 1.3cm,clip=true,height=4.5cm]{images/likelihood/lrew.pdf}
    \caption{Expected return as a function of the task likelihood.}
  \label{fig:likelihood}
\end{figure*}
